One exact and one random sampling method to assess the reliability of inferences from network enrichment analysis

### 2018-10-26 doc/genesetdp.tex
minor grammatic corrections, bibliography in the tex file

% As the often used network randomization techniques are not completely random, such methods often introduce an undesirable bias in the statistics. For instance, the actual set of random networks is much larger than the ones one use in such test, namely the random networks that retain a certain topological characteristic\cite{newman2001random}.
% Also, the somewhat inaccessible process used to randomize networks produces statistics that are equally hard to interpret, as such statistic in practice test a more complex null model than intended.
The methods that are used for network randomization are often mentally untractable, as they include a big number of steps that are subject to complex constraints.
As an example, in the link assignment plus second-order conservation method \cite{mccormack2013statistical}, links are assigned at random, but sequentially and at each step they must follow constraints that depend on previous link assignments.
While this results in a network that is both randomized and contains some desired topological features, preserved from the original network, these nested constraints make it very hard to verify that the resulting network is representative. In theory there is a space of all possible networks that could be made up from a certain number of nodes and node degree distribution, however, it is not clear that the link assignment and second-order conservation method samples evenly from that distribution.  And, if the resulting networks are not representative, the rejection of the resulting null hypothesis is uninformative, defeating the purpose of the statistical test.

@@

Hence, here we employ a different null model, more directly aimed at modelling the lack of association between query and pathway gene sets, which we formulate as ``there are not more links between the query and pathway gene sets than expected by chance''. We present a dynamic programming algorithm, which we dubbed GeneSetDP, that calculates the exact score distribution of any query of a given size. We also implemented one random sampling algorithm for the same task, that we called GeneSetMC. Both algorithms circumvent the need of parametric score distributions for the calculations of significance.
Combining the GeneSetDP and GeneSetMC with the scoring system of the popular method BinoX, we use simulations to demonstrate that both our algorithms produced unbiased statistics.

@@

\bibliography{./refs.bib}

>>>>

% As the often used network randomization techniques are not completely random, such methods often introduce an undesirable bias in the statistics. For instance, the actual set of random networks is much larger than the ones one use in such test, namely the random networks that retain a certain topological characteristic\cite{newman2001random}.
% Also, the somewhat inaccessible process used to randomize networks produces statistics that are equally hard to interpret, as such statistic in practice test a more complex null model than intended.
Yet, the methods that are used for network randomization are often mentally untractable, as they include a big number of steps that are subject to complex constraints.
As an example, in the link assignment plus second-order conservation method \cite{mccormack2013statistical}, links are assigned at random, but sequentially and at each step they must follow constraints that depend on previous link assignments.
While this results in a network that is both randomized and contains some desired topological features, preserved from the original network, these nested constraints make it very hard to verify that the resulting network is representative. In theory there is a space of all possible networks that could be made up from a certain number of nodes and node degree distribution, however, it is not clear that the link assignment and second-order conservation method samples evenly from that distribution.  And, if the resulting networks are not representative, the rejection of the resulting null hypothesis is uninformative, defeating the purpose of the statistical test.

@@

Hence, here we employ a different null model, more directly aimed at modelling the lack of association between query and pathway gene sets, which we formulate as ``there are not more links between the query and pathway gene sets than expected by chance''. We present a dynamic programming algorithm, which we dubbed GeneSetDP, that calculates the exact score distribution of any query of a given size. We also implemented one random sampling algorithm for the same task, that we called GeneSetMC. Both algorithms circumvent the need for parametric score distributions for the calculations of significance.
Combining the GeneSetDP and GeneSetMC with the scoring system of the popular method BinoX, we use simulations to demonstrate that both our algorithms produced unbiased statistics.

@@

\begin{thebibliography}{10}

\bibitem{tavazoie1999systematic}
Tavazoie S, Hughes JD, Campbell MJ, Cho RJ, Church GM.
\newblock Systematic determination of genetic network architecture.
\newblock Nature genetics. 1999;22(3):281.

\bibitem{khatri2012ten}
Khatri P, Sirota M, Butte AJ.
\newblock Ten years of pathway analysis: current approaches and outstanding
  challenges.
\newblock PLoS computational biology. 2012;8(2):e1002375.

\bibitem{goeman2007analyzing}
Goeman JJ, B\"{u}hlmann P.
\newblock Analyzing gene expression data in terms of gene sets: methodological
  issues.
\newblock Bioinformatics. 2007;23(8):980--987.

\bibitem{subramanian2005gene}
Subramanian A, Tamayo P, Mootha VK, Mukherjee S, Ebert BL, Gillette MA, et~al.
\newblock Gene set enrichment analysis: a knowledge-based approach for
  interpreting genome-wide expression profiles.
\newblock Proceedings of the National Academy of Sciences.
  2005;102(43):15545--15550.

\bibitem{bader2006pathguide}
Bader GD, Cary MP, Sander C.
\newblock Pathguide: a pathway resource list.
\newblock Nucleic acids research. 2006;34(suppl\_1):D504--D506.

\bibitem{szklarczyk2014string}
Szklarczyk D, Franceschini A, Wyder S, Forslund K, Heller D, Huerta-Cepas J,
  et~al.
\newblock STRING v10: protein--protein interaction networks, integrated over
  the tree of life.
\newblock Nucleic acids research. 2014;43(D1):D447--D452.

\bibitem{ogris2017funcoup}
Ogris C, Guala D, Kaduk M, Sonnhammer EL.
\newblock FunCoup 4: new species, data, and visualization.
\newblock Nucleic acids research. 2017;46(D1):D601--D607.

\bibitem{alexeyenko2012network}
Alexeyenko A, Lee W, Pernemalm M, Guegan J, Dessen P, Lazar V, et~al.
\newblock Network enrichment analysis: extension of gene-set enrichment
  analysis to gene networks.
\newblock {BMC} bioinformatics. 2012;13(1):226.

\bibitem{glaab2012enrichnet}
Glaab E, Baudot A, Krasnogor N, Schneider R, Valencia A.
\newblock EnrichNet: network-based gene set enrichment analysis.
\newblock Bioinformatics. 2012;28(18):i451--i457.

\bibitem{mccormack2013statistical}
McCormack T, Frings O, Alexeyenko A, Sonnhammer EL.
\newblock Statistical assessment of crosstalk enrichment between gene groups in
  biological networks.
\newblock {PloS} one. 2013;8(1):e54945.

\bibitem{ogris2016novel}
Ogris C, Guala D, Helleday T, Sonnhammer EL.
\newblock A novel method for crosstalk analysis of biological networks:
  improving accuracy of pathway annotation.
\newblock Nucleic acids research. 2016;45(2):e8--e8.

\bibitem{signorelli2016neat}
Signorelli M, Vinciotti V, Wit EC.
\newblock NEAT: an efficient network enrichment analysis test.
\newblock BMC bioinformatics. 2016;17(1):352.

\bibitem{lancaster1961significance}
Lancaster HO.
\newblock Significance tests in discrete distributions.
\newblock Journal of the American Statistical Association.
  1961;56(294):223--234.

\bibitem{hwang2001optimality}
Hwang JG, Yang MC.
\newblock An optimality theory for mid p--values In 2$\times$ 2 contingency
  tables.
\newblock Statistica Sinica. 2001; p. 807--826.

\end{thebibliography}
